R package \texttt{glmm}

### 2014-10-20 CompareExamples/glmmexamples.tex
working on CLT proof for gradient

\title{R package glmm}

@@

\section{Bacteria example}
The MASS package contains the command glmmPQL and the bacteria data-set. The manual describes the data set as follows: ``Tests of the presence of the bacteria H. influenzae in children with otitis media in the Northern Territory of Australia.''

The data were fit using glmmPQL, glmm with $m=10^3$, glmm with $m=10^4$, glmm with $m=10^5$. (The data did need to be reformatted, which took only a couple minutes). The parameter estimates are summarized in the following table. More model details can be seen in the output that follows the table.

\begin{tabular}{lccccc}
& Intercept & trtdrug & trtdrug+ & I(week$>2$) TRUE & $\nu$ \\ \hline
glmmPQL & 3.41 & -1.25 & -.75 & -1.61 & 1.99\\
glmm $m=10^3$ & 3.02 & -1.20 & -.89 & -1.44 & .81 \\
glmm $m=10^4$ & 3.65 & -1.36 & -.92 & -1.66 & 2.02 \\
glmm $m=10^5$ & 3.49& -1.65 & -.90 & -1.46 & .90 \\
\end{tabular}

It's safe to say that the bacteria glmm results with a paltry Monte Carlo sample size of $m=10^3$ are not reliable. We conclude this because the estimates change quite a bit when $m=10^4$. The results using $m=10^4$ are very similar to the glmmPQL results.

The glmmPQL results:
\begin{verbatim}
> bac.pql<-glmmPQL(y ~ trt + I(week > 2), random = ~ 1 | ID,
+                 family = binomial, data = bacteria)
> summary(bac.pql)
Linear mixed-effects model fit by maximum likelihood
 Data: bacteria 
  AIC BIC logLik
   NA  NA     NA

@@

The bacteria glmm results with a  Monte Carlo sample size of $m=10^3$:
\begin{verbatim}
> set.seed(1234)
> bac.glmm1<-glmm(y2~trt+I(week > 2),list(~0+ID), 
family=bernoulli.glmm, data=bacteria, m=10^3, varcomps.names=c("ID"))

> summary(bac.glmm1)

Call:
glmm(fixed = y2 ~ trt + I(week > 2), random = list(~0 + ID),  varcomps.names = c("ID"), 
data = bacteria, family.glmm = bernoulli.glmm,     m = 10^3)

Fixed Effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.0199     0.4667   6.471 9.75e-11 ***
trtdrug          -1.1989     0.4473  -2.680 0.007357 ** 
trtdrug+         -0.8883     0.4742  -1.873 0.061021 .  
I(week > 2)TRUE  -1.4407     0.4258  -3.384 0.000716 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
   Estimate Std. Error z value Pr(>|z|)/2    
ID   0.8175     0.1635   4.998   2.89e-07 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

\end{verbatim}
The results from fitting the bacteria dataset with glmm and $m=10^4$:

\begin{verbatim}
> set.seed(1234)
> bac.glmm2<-glmm(y2~trt+I(week > 2),list(~0+ID), family=bernoulli.glmm, data=bacteria, m=10^4, varcomps.names=c("ID"))
> summary(bac.glmm2)

@@

Fixed Effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.6514     0.6327   5.771  7.9e-09 ***
trtdrug          -1.3645     0.7461  -1.829 0.067427 .  
trtdrug+         -0.9186     0.7619  -1.206 0.227915    
I(week > 2)TRUE  -1.6660     0.4645  -3.587 0.000334 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
   Estimate Std. Error z value Pr(>|z|)/2   
ID   2.0244     0.7515   2.694    0.00353 **
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

@@

Results  of glmm with $m=10^5$:
\begin{verbatim}
> set.seed(1234)
> bac.glmm3<-glmm(y2~trt+I(week > 2),list(~0+ID), family=bernoulli.glmm, data=bacteria, m=10^5, varcomps.names=c("ID"))
> summary(bac.glmm3)

@@

Fixed Effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.4888     0.4861   7.178 7.08e-13 ***
trtdrug          -1.6526     0.5232  -3.159 0.001586 ** 
trtdrug+         -0.8968     0.5160  -1.738 0.082196 .  
I(week > 2)TRUE  -1.4614     0.4334  -3.372 0.000746 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
   Estimate Std. Error z value Pr(>|z|)/2    
ID   0.8999     0.2471   3.641   0.000136 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

@@

@@

First, I fit the data using glmer in lme4. The model summary says it was fit with Adaptive Hermite Quadrature with 9 quadrature nodes. Then I fit the data using glmm with $m=10^4$. This took 16.75 minutes on my netbook. The point estimates and some of the standard errors are very similar between the two  methods of model-fitting.

\begin{tabular}{lccccc}
& Intercept & period2 & period3 & period4 & $\nu$ \\ \hline
glmer (numerical integration) & -1.40 & -.99 & -1.13 & -1.58 & .41\\
glmm $m=10^4$ & -1.42 & -.99 & -1.13 & -1.58 & .44 \\
\end{tabular}

@@

Next, the results using glmm with $m=10^4$:
\begin{verbatim}
> summary(herd.glmm1)

@@

I fit the salamander data set using  my glmm package (with $m=10^4$ twice and with $m=10^6$ once)  and compared those results to the MCEM results of Booth and Hobert. They do not look so great, and the $m=10^6$ results took 41 hours to get.

\begin{tabular}{lcccccc}
& Intercept & RW & WR & WW & $\nu_F$ & $\nu_M$ \\ \hline
MCEM (B+H) & 1.03 & .32 & -1.95 & .99 & 1.4 & 1.25 \\
glmm $m=10^5$ &   0.964      -0.645      -2.767      -0.008
\end{tabular}

\section{Murder}
Subjects were asked how many victims of homicide they personally knew. The data set is from Agresti's \textit{Categorical Data Analysis} book. I found similar answers from my glmm package as from glmmPQL.

@@

The murder model resulting from glmmPQL:
\begin{verbatim}
> murder.pql<-glmmPQL(y~race,random=~1|race,data=murder,family=poisson)iteration 1
> summary(murder.pql)
Linear mixed-effects model fit by maximum likelihood
 Data: murder 
  AIC BIC logLik
   NA  NA     NA

>>>>

\title{R package \texttt{glmm}}

@@

\section{Bacteria example}
The MASS package contains the command \texttt{glmmPQL} and the bacteria data-set. The manual describes the data set as follows: ``Tests of the presence of the bacteria H. influenzae in children with otitis media in the Northern Territory of Australia.''

The data were fit using \texttt{glmmPQL},  \texttt{glmer}  from the \texttt{lme4} package (method implemented was Laplace approximation),  \texttt{glmm} with $m=10^4$,  \texttt{glmm} with $m=10^5$, and \texttt{glmm} with $m=10^6$. (The data did need to be reformatted in a process described at the end of the document). The parameter estimates are summarized in the following table. More model details can be seen in the output that follows the table.

\begin{tabular}{lccccc}
& Intercept & trtdrug & trtdrug+ & I(week$>2$) TRUE & $\nu$ \\ \hline
\texttt{glmmPQL} & 3.41 & -1.25 & -.75 & -1.61 & 1.99\\
\texttt{glmer} & 3.55 & -1.37 & -.78 & -1.60 & 1.54\\
\texttt{glmm} $m=10^4$ & 3.57 & -1.38 & -.80 & -1.62 & 1.66 \\
\texttt{glmm} $m=10^5$ & 3.60& -1.39 & -.79 & -1.63 & 1.73 \\
\texttt{glmm} $m=10^6$ & 3.58& -1.37 & -.79 & -1.63 & 1.71 \\
\end{tabular}

The \texttt{glmmPQL} results:
\begin{verbatim}
> bac.pql<-glmmPQL(y ~ trt + I(week > 2), random = ~ 1 | ID,
+                 family = binomial, data = bacteria)
> summary(bac.pql)
Linear mixed-effects model fit by maximum likelihood
 Data: bacteria 
  AIC BIC logLik
   NA  NA     NA

@@

\begin{verbatim}
proc.time()-ptm
user  system elapsed 
 16.226   0.044  16.281 
> summary(bac.glmm2)

@@

Fixed Effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.5745     0.6371   5.611 2.01e-08 ***
trtdrug          -1.3779     0.6701  -2.056 0.039768 *  
trtdrug+         -0.8014     0.6878  -1.165 0.243950    
I(week > 2)TRUE  -1.6224     0.4674  -3.471 0.000518 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
   Estimate Std. Error z value Pr(>|z|)/2  
ID   1.6635     0.7807   2.131     0.0165 *
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

@@

Results from  \texttt{glmm}  with $m=10^5$.
\begin{verbatim}
> bac.glmm2<-glmm(y2~trt+I(week > 2),list(~0+ID), family=bernoulli.glmm, data=bacteria, m=10^5, varcomps.names=c("ID"))
> proc.time()-ptm
   user  system elapsed 
175.729   0.150 176.211 
> summary(bac.glmm2)

@@

Fixed Effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.6008     0.7666   4.697 2.64e-06 ***
trtdrug          -1.3868     0.7126  -1.946 0.051641 .  
trtdrug+         -0.7940     0.6996  -1.135 0.256348    
I(week > 2)TRUE  -1.6321     0.4917  -3.319 0.000903 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
   Estimate Std. Error z value Pr(>|z|)/2  
ID    1.738      1.317    1.32     0.0935 .
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

@@

Results from \texttt{glmm} with $m=10^6$.
\begin{verbatim}
 proc.time()-ptm
    user   system  elapsed 
1491.511    1.299 1494.133 
> summary(bac.glmm2)

Call:
glmm(fixed = y2 ~ trt + I(week > 2), random = list(~0 + ID), 
    varcomps.names = c("ID"), data = bacteria, family.glmm = bernoulli.glmm, 
    m = 10^6)

Fixed Effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.5824     0.6952   5.153 2.57e-07 ***
trtdrug          -1.3714     0.6935  -1.978 0.047980 *  
trtdrug+         -0.7929     0.7034  -1.127 0.259644    
I(week > 2)TRUE  -1.6278     0.4801  -3.391 0.000698 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Variance Components for Random Effects (P-values are one-tailed):
   Estimate Std. Error z value Pr(>|z|)/2  
ID    1.707      1.050   1.626      0.052 .
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
\end{verbatim}

Results from \texttt{lme4}.
\begin{verbatim}
 summary(bac.glmer)
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial ( logit )
Formula: y2 ~ trt + I(week > 2) + (1 | ID)
   Data: bacteria

     AIC      BIC   logLik deviance df.resid 
   202.3    219.2    -96.1    192.3      215

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-4.5615  0.1359  0.3022  0.4217  1.1276

Random effects:
 Groups Name        Variance Std.Dev.
 ID     (Intercept) 1.543    1.242   
Number of obs: 220, groups: ID, 50

Fixed effects:
                Estimate Std. Error z value Pr(>|z|)    
(Intercept)       3.5479     0.6958   5.099 3.41e-07 ***
trtdrug          -1.3667     0.6770  -2.019 0.043517 *  
trtdrug+         -0.7826     0.6831  -1.146 0.251926    
I(week > 2)TRUE  -1.5985     0.4759  -3.359 0.000783 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Correlation of Fixed Effects:
            (Intr) trtdrg trtdr+
trtdrug     -0.593              
trtdrug+    -0.537  0.487       
I(wk>2)TRUE -0.656  0.126  0.064
> 
\end{verbatim}

@@

First, I fit the data using glmer in lme4. The model summary says it was fit with Adaptive Hermite Quadrature with 9 quadrature nodes. Then I fit the data using \texttt{glmm}with $m=10^4$. This took 16.75 minutes on my netbook. The point estimates and some of the standard errors are very similar between the two  methods of model-fitting.

\begin{tabular}{lccccc}
& Intercept & period2 & period3 & period4 & $\nu$ \\ \hline
glmer (numerical integration) & -1.40 & -.99 & -1.13 & -1.58 & .41\\
\texttt{glmm}$m=10^4$ & -1.42 & -.99 & -1.13 & -1.58 & .44 \\
\end{tabular}

@@

Next, the results using \texttt{glmm}with $m=10^4$:
\begin{verbatim}
> summary(herd.glmm1)

@@

I fit the salamander data set using  my \texttt{glmm} package (with $m=10^4$ twice and with $m=10^6$ once)  and compared those results to the MCEM results of Booth and Hobert. They do not look so great, and the $m=10^6$ results took 41 hours to get.

\begin{tabular}{lcccccc}
& Intercept & RW & WR & WW & $\nu_F$ & $\nu_M$ \\ \hline
MCEM (B+H) & 1.03 & .32 & -1.95 & .99 & 1.4 & 1.25 \\
\texttt{glmm} $m=10^5$ &   0.964      -0.645      -2.767      -0.008
\end{tabular}

\section{Murder}
Subjects were asked how many victims of homicide they personally knew. The data set is from Agresti's \textit{Categorical Data Analysis} book. I found similar answers from my glmm package as from \texttt{glmmPQL}.

@@

The murder model resulting from \texttt{glmmPQL}:
\begin{verbatim}
> murder.pql<-glmmPQL(y~race,random=~1|race,data=murder,family=poisson)iteration 1
> summary(murder.pql)
Linear mixed-effects model fit by maximum likelihood
 Data: murder 
  AIC BIC logLik
   NA  NA     NA
